---
title: "SimulationReport"
author: "bbneo"
date: "03/18/2015"
output: pdf_document
---

    
# Simulation Exercise:  

The distribution of a sample mean statistic with reference to the population mean and standard deviation and its relation to the `t` and `Z` distributions.

## Overview
```{r, echo=FALSE}
n = 40
lambda = 0.2
set.seed(3.1415926)
options(digits=3)
```

In this report, we will illustrate the nature of the distribution of the `n = ` `r n` *sample mean* statistic from an exponential population distribution.  What we will see is that:

- the *sample mean* statistic can be a good estimator of the mean of the population (exponential) distribution,
- the distribution of the sample means approaches a `t` or `Z` distribution as the sample size increases as expected from the Central Limit Theorem,
- the `standard` *`error`* of this population mean estimator (`SE`_`mean`_) depends on the `standard deviation` of the sample means and the sample size, and
- the `standard error` of this *mean estimator* has a much narrower confidence interval for the **`population mean`** estimate than either
    - the *standard deviation* of the distribution of the sample means, or
    - the *standard deviation* of the population distribution itself


## The `exponential` distribution

The population distribution we will use for this illustration is the *exponential* distribution.  The `exponential` distribution is implemented in R as `dexp/rexp`.  

The distribution has one parameter, the `rate` or `lambda`, and both the theoretical `mean` **and** `standard deviation` (of random *single* samples) of the (*population*) distribution are  `1 / lambda`.

For `lambda = ` `r lambda`, `rexp` looks like this:

```{r, echo=FALSE}
xMax <- 1/lambda + 2 * 1/lambda
theorMean <- 1/lambda; theorSd <- 1/lambda

x <- seq(0,xMax,1/lambda/50)
dist <- dexp(x,rate=lambda)

plot(x, dist, main=paste("The exponential distribution, lambda =", lambda), type = 'l')
abline(v = theorMean, lwd = 2)
abline(v = theorMean + c(-1,1)*theorSd, lwd = 2, lty = "dashed")
```

The population `mean` is represented as solid vertical line at `r theorMean`, and the dashed lines represent `+/- 1 standard deviation` of the distribution at `r theorMean + c(-1,1)*theorSd`.  Notice that this distribution is skewed and does not have a tail approaching zero on the left.  

## Simulation and the *sample mean* as an *estimator* of the *population mean*

We will take `1000` **\*** (`n = ` `r n`) samples from the popultation distribution `rexp`, and use these samples to estimate the *population mean* and the *standard error* of that *mean* estimate (`SE`_`mean`_).  We create these samples in R by successively adding `n = ` `r n` *sample means* to a vector `mns`.

```{r}
mns <- NULL
for (i in 1:1000) mns <- c(mns, mean(rexp(n,lambda)) )
```

## The distribution of the *sample means*

The distribution of these `sample means` approaches a `t` or `Z` distribution as is shown in the figure below:


```{r, echo=FALSE}
sampleMean <- mean(mns)   # mean of the sample means
sampleSd <- sd(mns)       # sd of the sample means
scale <- 200/0.4

hist(mns,  xlim = c(0,10), ylim = c(0,250),
     main = paste("Histogram of Means, 1000 * n =", n, 
                  "Samples rexp, lambda=",lambda),
     xlab="Sample means")

abline(v=theorMean, lwd=2)
abline(v=(theorMean + c(-1,1)*theorSd), lwd=2, lty="dashed")

abline(v=sampleMean,lwd=2,col="darkblue")
abline(v=sampleMean + c(-1,1)*qnorm(0.975)*sampleSd,lwd=2,col="darkblue",lty="dotted")

# overlay the t and normal distributions

lines(x,scale*dt((x-sampleMean),df=n-1), col = "darkgreen", lwd = 2)
lines(x,scale*dnorm(x,mean=sampleMean,sd=sampleSd), col = "darkred", lwd = 2)
```

The `mean` of the *sample means* is `r sampleMean` (represented as a solid vertical dark blue line) with a 95% confidence interval for the distribution of the *sample means* represented by the dotted dark blue vertical lines on the figure at:

```{r} 
sampleMean + c(-1,1)*qnorm(0.975)*sampleSd
```  


The *population mean* and standard deviation are represented by a vertical line at `x = ` `r theorMean`, and dashed lines at `x = ` `r theorMean + c(-1,1)*theorSd`.  The population distribution for **single** random iid samples from an exponential distribution would give the very wide `95%` *standard* quartile `r c(-1,1)*qnorm(0.975)` `* theoreticalSd` interval of **`r theorMean + c(-1,1)*qnorm(0.975)*theorSd` !!**  

But the skewness of the `exponential` distribution makes the normal distribution a poor tool for describing its variation.

A `t density distribution` with `df= ` `r n-1` is overlaid in dark green.

A `normal` density distribution with mean and sd corresponding to the sample distribution is overlaid in dark red.

## A closer look at the confidence interval for the population mean estimate

We will zoom in on the 95% confidence interval of the *sample mean* and its `SE`_`mean`_ as an estimate of the *population mean* of `rexp`.

```{r, echo=FALSE}
hist(mns,  xlim = c(4,6), ylim = c(0,250),
     main = paste("Histogram of Means, 1000 * n =", n, 
                  "Samples rexp, lambda=",lambda),
     xlab="Sample means")

abline( v = theorMean, lwd = 4 )

abline(v = sampleMean, col = "darkblue", lwd = 4, lty = "dotted")
abline(v=sampleMean + c(-1,1)*qnorm(0.975)*sampleSd,lwd=2,col="darkblue",lty="dotted")

abline( v = c(sampleMean + c(-1,1)*qt(0.975,n-1)*sampleSd/sqrt(n)), 
        col = "red", lwd = 2, lty = "longdash" )

lines(x, scale*dt((x-sampleMean),df=n-1), 
      col = "darkgreen", lwd = 2)

lines(x, scale*dnorm(x,mean=sampleMean,sd=sampleSd), 
      col = "darkred", lwd = 2)
```

The 95% confidence interval for the estimation of the `population (theoretical) mean` from this set of samples using a `t` statistic is:  

```{r, collapse=TRUE}
SEmean <- sampleSd/sqrt(n)
SEmean

qt(0.975,n-1)
sampleMean + c(-1,1)*qt(0.975,n-1)*SEmean
```

which is represented in the figure by vertical red dashed lines.  

The 95% CI for the *sample mean* as an estimator of the *population mean* is (using a Z quartile):

```{r, collapse=TRUE}
qnorm(0.975)
sampleMean + c(-1,1)*qnorm(0.975)*SEmean
```

Which is very close to the 95% CI obtained using a `t` statistic, just slightly more narrow.

\pagebreak

## In Summary

For this report, for `n =` `r n` `> 20-30`, the distribution of sample means approaches a normal `Z` distribution:

statistic | mean | sd | 95% CI 
----------|------|-----|--------
population theoretical | `r theorMean` | `r theorSd` | --
dist. of sample mean | `r sampleMean` | `r sampleSd` | `r sampleMean + c(-1,1)*qnorm(0.975)*sampleSd`
pop. mean *estim* | `r sampleMean` |  | `r sampleMean + c(-1,1)*qnorm(0.975)*SEmean`


- the *sample mean* statistic can be a good estimator of the mean of the population (exponential) distribution,

- the distribution of the sample means approaches a `t` or `Z` distribution as the sample size increases as expected from the Central Limit Theorem,

- the `standard` *`error`* of this population mean estimator (`SE`_`mean`_) depends on the `standard deviation` of the sample means and the sample size, and

- the `standard error` of this *mean estimator* has a much narrower confidence interval for the **`population mean`** estimate than either:

    - the *standard deviation* of the distribution of the *sample* means, or
    - the *standard deviation* of the *population distribution itself* (the distribution of a large collection of exponentially distributed random numbers)

